---
title: "hw2"
author: "Zihao_Wang"
date: "11/3/2018"
output: html_document
---
```{r}
rm(list=ls())
set.seed(12345)
options(warn = -1)
knitr::opts_knit$set(root.dir = '~/Desktop/finm331-fall-2018/analysis/')
library(ggplot2)
```

# 1
```{r}
track = read.table("../data/hw2/p1.txt")
colnames(track) = c("Country","100m","200m","400m","800m","1500m","3000m","Marathon")
#track$Country = as.factor(track$Country)
X = track[,colnames(track) != "Country"]
X = as.matrix(X)
n = nrow(X)
p = ncol(X)

## sample means
### of course the sample means for "Country" variable 
Xbar = (t(replicate(n,1)) %*% X)/n

## sample covariance and correlation matrix
S = (t(X) %*% (diag(replicate(n,1)) - matrix(replicate(n*n,1), nrow = n)/n)  %*% X)/(n-1)

D = diag(1/sqrt(diag(S)))
R =  D %*% S %*% D

## correlation matrix of the log of data
X_log = log(X)
S_log = (t(X_log) %*% (diag(replicate(n,1)) - matrix(replicate(n*n,1), nrow = n)/n)  %*% X_log)/(n-1)
D_log = diag(1/sqrt(diag(S_log)))
R_log = D_log %*% S_log %*% D_log

format(Xbar,digits = 2)
format(S,digits = 2)
format(R,digits = 2)
format(R_log,digits = 2)
```

## Comment
It is surprising that the correlation matrix of the log data is almost the same as that of the original data.

# 4 (Sample PCA)
## (a)
```{r}
eigdec = eigen(R)
print(paste0("sum of eigenvalues: ", sum(eigdec$values), " while the number of feature p: ",p))
```
### Comment
The sum of eigenvalues of sample correlation matrix is the same as the $p$, the number of features.

## (b)

### 1
```{r}
## standardize X (call it Z)
Z = (X - t(replicate(n,as.numeric(Xbar)))) %*% D
## svd
svd_Z = svd(Z)
pc1 = svd_Z$v[,1]
q1 = eigdec$vectors[,1]
print("ratio of pc1 from svd to pc1 from evd is: ")
pc1/q1
```

### 2
Use the formula $\lambda_k = \frac{1}{n-1} \sigma_k^2$, we know the proportion is $\frac{\lambda_k}{\lambda_1+...+\lambda_p} = \frac{\sigma_k^2}{\sigma_1^2+...+\sigma_p^2}$. So the ratio explained by the first two pcs is:
```{r}
sum(svd_Z$d[1:2]^2)/sum(svd_Z$d^2)
```

## (c)

### 1
```{r}
Score = svd_Z$u[,1:2]
Score = as.data.frame(Score)
Score = cbind(as.factor(track$Country),Score)
names(Score) <- c("Country","PC1","PC2")
ggplot(Score, aes(x=PC1, y = PC2, label = Country))+
  geom_point(color = "blue")+
  geom_text()
```

### 2
```{r}
Score = Score[order(Score$PC1),]
print("top 6 in PC1: ")
head(Score$Country,6)

print("last 3 in PC1: ")
tail(Score$Country,3)
```


#### Comment:
The list is similar to the rank of the number of medals won in the Olympic Games. "Athletic Excellence" is a very subjective definition. If I define it in terms of the number of top prizes won by a country, then it reflects "Athletic Excellence".

# 5 (Scaling effect in sample PCA)
```{r}
#rm(list = ls())
air = read.table("../data/hw2/p5.txt")
colnames(air) = c("Wind", "Radiation", "CO", "NO", "NO2", "O3", "HC")

n = nrow(air)
p = ncol(air)
```

## (a)
```{r}
X = as.matrix(air)
Xbar = as.numeric((t(replicate(n,1)) %*% X) / n)
Xc = X - t(replicate(n,Xbar))

## PCA on sample covariance matrix
svd_Xc = svd(Xc)

## Scree plot (well, not quite the same)
idx = 1:7
scree1 = sapply(idx, function(i) sum(svd_Xc$d[1:i]))/sum(svd_Xc$d)
plot(idx,scree1)

```

```{r}
## standardize data
S = (t(X) %*% (diag(replicate(n,1)) - matrix(replicate(n*n,1), nrow = n)/n)  %*% X)/(n-1)
D = diag(1/sqrt(diag(S)))
Z = Xc %*% D

## PCA on sample covariance matrix
svd_Z = svd(Z)

## Scree plot (well, not quite the same)
idx = 1:7
scree2 = sapply(idx, function(i) sum(svd_Z$d[1:i]))/sum(svd_Z$d)
plot(idx,scree2)

```
### Comment:
* In the case of covariance matrix, we need 4 PCs to account for 90% of the variance; \

* In the case of correlation matrix, we need 6 PCs to account for 90% of the variance

## (b)
```{r}
plot(svd_Xc$d/sum(svd_Xc$d), ylab = "variance explained", xlab = "PC", main = "Sample Covariance ")
plot(svd_Z$d/sum(svd_Z$d), ylab = "variance explained", xlab = "PC", main = "Sample Correlation ")

```
## (c)
Comment: Without scaling, the variance can be explained by a smaller number of PCs, as opposed to the scaled data.


# 6
```{r}
#rm(list = ls())
data <- read.csv("../data/hw2/p6-data.txt", sep = ",", row.names = 1,header = TRUE)
```

## (a) standardizing
```{r}
std_data <- function(X){
  n = nrow(X)
  p = ncol(X)
  Xbar = as.numeric((t(replicate(n,1)) %*% X) / n)
  Xc = X - t(replicate(n,Xbar))
  S = (t(X) %*% (diag(replicate(n,1)) - matrix(replicate(n*n,1), nrow = n)/n)  %*% X)/(n-1)
  D = diag(1/sqrt(diag(S)))
  Z = Xc %*% D 
  return(Z)
}

A = as.matrix(data)
Ahat = std_data(A)
```

## (b) SVD + PCA scatter plot
```{r}
svd_Ahat = svd(Ahat)
Score = svd_Ahat$u %*% diag(svd_Ahat$d)
Score = as.data.frame(Score[,1:2])
Score = cbind(as.factor(rownames(data)),Score)
names(Score) <- c("Country_Region","PC1","PC2")

p1 = ggplot(Score, aes(x=PC1, y = PC2, label = Country_Region))+
  geom_point(color = "blue")+
  geom_text()
p1

```
### Comment:
There is a common linear relation between PC1 and PC2 for most of the countries, except the two outliers : Hong Kong and Singapore

## (c) SVD + variable PCA scatter plot
```{r}
Score_var = svd_Ahat$v %*% diag(svd_Ahat$d)
Score_var = as.data.frame(Score_var[,1:2])
Score_var = cbind(as.factor(colnames(data)),Score_var)
names(Score_var) <- c("feature","vPC1","vPC2")

p2 = ggplot(Score_var, aes(x=vPC1, y = vPC2, label = feature))+
  geom_point(color = "purple")+
  geom_text()
p2

```

## (d) Biplot
```{r}
p1 = ggplot(Score, aes(x=PC1, y = PC2, label = Country_Region))+
  geom_point(color = "black")+
  geom_text(color = "blue")

p2 = p1 +
  geom_point(data = Score_var, aes(x=vPC1,y = vPC2, label = feature), color = "red")+
  geom_text(data = Score_var, aes(x=vPC1,y = vPC2, label = feature), color = "green")

p2

```

### Comment:
From the biplot, we can see that HK and Singapore are outliers because of the two features, the 
"Population.sq.km" and "Population.1000HectarAgri".

## (e) Outlier removed, redo (a), (b)
```{r}
A_rm = A[! rownames(A) %in% c("Hong Kong","Singapore"),]
Ahat = std_data(A_rm)

svd_Ahat = svd(Ahat)
Score = svd_Ahat$u %*% diag(svd_Ahat$d)
Score = as.data.frame(Score[,1:2])
Score = cbind(as.factor(rownames(A_rm)),Score)
names(Score) <- c("Country_Region","PC1","PC2")

p1 = ggplot(Score, aes(x=PC1, y = PC2, label = Country_Region))+
  geom_point(color = "blue")+
  geom_text()
p1

```
### Comment:
Belgium and netherlands are the two European countries closest to Japan.

  








